---
title: "vgscKO larval survival"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
library(dplyr); packageVersion("dplyr")
library(lme4); packageVersion("lme4")
library(ggplot2); packageVersion("ggplot2")
library(arm); packageVersion("arm")
library(emmeans); packageVersion("emmeans")
library(DHARMa); packageVersion("DHARMa")
```

Load data

```{r}
df <- read.csv("SurvivalRate.csv")
head(df)
```

## Binomial GLMM with complementary log-log link
First reformat data to compute at risk numbers (necessary for the GLMM)
```{r}
cp <- df %>%
  # rename columns
  rename(
    genotype  = Genotype,
    tray_raw  = Tray,
    tray_size = NoLarveStart,
    day       = Day,
    alive     = Larvae
  ) %>%
  
  # Create unique tray IDs (since A, B, C appear in each genotype)
  mutate(
    genotype  = factor(genotype),
    tray_size = factor(tray_size, levels = c("25L", "50L")),
    day       = as.integer(day),
    tray_id   = interaction(genotype, tray_raw, drop = TRUE, sep = "_")
  ) %>%
  
  arrange(genotype, tray_id, day) %>%
  group_by(genotype, tray_id, tray_size) %>%
  
  # Compute alive_next and deaths
  mutate(
    alive_next = lead(alive),

    # For the last observed day:
    # larvae are censored (so alive_next = alive → deaths = 0)
    alive_next = ifelse(is.na(alive_next), alive, alive_next),

    # Number dying in the interval (day to day+1)
    deaths = pmax(alive - alive_next, 0L),

    # Number at risk at the start of the day
    at_risk = alive
  ) %>%
  ungroup()

cp <- cp %>%
  mutate(genotype = relevel(genotype, ref = "G3"))  # sets G3 as control
```

Binomial GLMM

```{r}
m1 <- glmer(
  cbind(deaths, at_risk - deaths) ~ genotype + tray_size + factor(day) + (1 | tray_id),
  data = cp,
  family = binomial(link = "cloglog"),
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))
)

summary(m1)
```

Model checks - whether genotype impacts the model, whether there are interactions between tray and 
genotype, and overdispersion checks

```{r}
# Comparing models with and without genotype (to see if genotype makes a difference)
m0 <- update(m1, . ~ . - genotype)
anova(m0, m1, test = "Chisq")
```

```{r}
# Check for overdispersion
rdf <- df.residual(m1)
pearson <- sum(residuals(m1, type = "pearson")^2)
overdispersion_ratio <- pearson / rdf
overdispersion_ratio
```

Check residuals (should be <2 for good fit)

```{r}
m1_Model_DevResid <- resid (m1, type = "deviance")
hist(m1_Model_DevResid)
```

Check variance of residuals across the range of fitted values (most points should fall within the grey line)

```{r}
m1_Model_Predict <- predict(m1)
binnedplot(m1_Model_Predict, (m1_Model_DevResid))
```

Check for outliers (Cook's distance >1 indicates outliers)
```{r}
plot(m1, which = 4)
```

Double check fit with DHARMa

```{r}
sim <- simulateResiduals(m1)
plot(sim)
```
```{r}
testDispersion(sim)
testUniformity(sim)
testOutliers(sim)
```
```{r}
plotResiduals(sim, cp$genotype)
plotResiduals(sim, cp$tray_size)
plotResiduals(sim, cp$day)
```

Quantile deviations observed in day, potentially because day is treated as a factor
when there is next to no observable death on two of the days. Could try a simpler model
as random effects have no impact on variance.

```{r}
m1_simpler <- glm(
  cbind(deaths, at_risk - deaths) ~ genotype + tray_size + factor(day),
  data = cp,
  family = binomial(link = "cloglog")
)
summary(m1_simpler)
```


```{r}
# Check for overdispersion
rdf <- df.residual(m1_simpler)
pearson <- sum(residuals(m1_simpler, type = "pearson")^2)
overdispersion_ratio <- pearson / rdf
overdispersion_ratio
```

Check model fit with DHARMa

```{r}
sim_simpler <- simulateResiduals(m1_simpler)
plot(sim_simpler)
```
```{r}
testDispersion(sim_simpler)
testUniformity(sim_simpler)
testOutliers(sim_simpler)
```
```{r}
plotResiduals(sim_simpler, cp$genotype)
plotResiduals(sim_simpler, cp$tray_size)
plotResiduals(sim_simpler, cp$day)
```

No flagged issues, simpler model seems to fit better.


Calculate hazard ratios based on simple model

```{r}
coefs <- summary(m1_simpler)$coefficients
hr <- data.frame(
  term  = rownames(coefs),
  HR    = exp(coefs[, "Estimate"]),
  lower = exp(coefs[, "Estimate"] - 1.96 * coefs[, "Std. Error"]),
  upper = exp(coefs[, "Estimate"] + 1.96 * coefs[, "Std. Error"]),
  p     = coefs[, "Pr(>|z|)"]
)

hr[grep("genotype|tray_size", hr$term), ]
```

**Interpretation**

No significant difference in survival between the three genotypes

## Survival predictions at population level, and plotting

As tray size or individual tray had no significant impact in the model, hazard ratios are pooled 
for trays in each genotype and day

```{r}
# Create a prediction grid for each genotype and day
newd <- expand.grid(
  genotype = levels(cp$genotype),
  tray_size = levels(cp$tray_size),
  day = 1:10
)

newd$day <- factor(newd$day)

# Convert cloglog(eta) to probability of death (hazard) for each day
inv_cloglog <- function(eta) 1 - exp(-exp(eta))

# Predict hazard (population level)
eta_hat <- predict(
  m1_simpler,
  newdata = newd,
  type = "link",
  re.form = NA
)
newd$hazard_hat <- inv_cloglog(eta_hat)

# Average hazards across tray sizes for each genotype × day
haz_by_gen_day <- newd %>%
  group_by(genotype, day) %>%
  summarise(hazard = mean(hazard_hat), .groups = "drop") %>%
  arrange(genotype, as.integer(as.character(day))) %>%
  group_by(genotype) %>%
  mutate(Survival = cumprod(1 - hazard)) %>%
  ungroup() %>%
  mutate(day = as.integer(as.character(day)))

```

Bootstrapping for 95% CIs (x1000)
```{r}
stat_fun <- function(fit) {
  eta <- predict(fit, newdata = newd, type = "link", re.form = NA)
  haz <- inv_cloglog(eta)
  tmp <- newd
  tmp$hazard <- haz
  
  out <- tmp %>%
    group_by(genotype, day) %>%
    summarise(hazard = mean(hazard), .groups = "drop") %>%
    arrange(genotype, as.integer(as.character(day))) %>%
    group_by(genotype) %>%
    mutate(Survival = cumprod(1 - hazard)) %>%
    ungroup() %>%
    mutate(day = as.integer(as.character(day)))
  out$Survival
}

set.seed(202)
nsim <- 1000
bt <- bootMer(m1, FUN = stat_fun, nsim = nsim, use.u = FALSE,
              type = "parametric", parallel = "no")

G <- length(levels(cp$genotype))
D <- 10
stopifnot(ncol(bt$t) == G * D)

ci_list <- vector("list", G)
gen_levels <- levels(cp$genotype)

for (g in seq_len(G)) {
  cols <- ((g - 1) * D + 1):(g * D)
  qlo <- apply(bt$t[, cols, drop = FALSE], 2, quantile, probs = 0.05, na.rm = TRUE)
  qhi <- apply(bt$t[, cols, drop = FALSE], 2, quantile, probs = 0.95, na.rm = TRUE)
  ci_list[[g]] <- data.frame(
    genotype = gen_levels[g],
    day = 1:D,
    lo = qlo,
    hi = qhi
  )
}

ci_df <- bind_rows(ci_list)

plot_df <- haz_by_gen_day %>%
  left_join(ci_df, by = c("genotype", "day"))
write.csv(plot_df, "SurvivalPlotdf.csv")
```

Plotting prediction curve with 95% CI ribbon

```{r}
genotype_cols2 <- c(
  "CFP+" = "#0B3C8C",
  "CFP-" = "#B11226",
  "G3"   = "#0E7C3A"
)

genotype_cols3 <- c(
  "CFP+" = "#0c4084",
  "CFP-" = "#ac1925",
  "G3"   = "#19a453"
)

genotype_names <- c(
  "CFP+" = expression(vgscKO^"+"),
  "CFP-" = expression(vgscKO^"-"),
  "G3"   = expression(G3)
)

ggplot(plot_df, aes(x = day, y = Survival, color = genotype, fill = genotype)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1.5) +
  scale_color_manual(values = genotype_cols3, labels = genotype_names) +
  scale_fill_manual(values = genotype_cols2, labels = genotype_names) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0.8,1)) +
  scale_x_continuous(breaks = 1:10) +
  labs(
    x = "Day",
    y = "Predicted survival",
  ) +
  theme_classic() +
  theme(legend.position = "bottom") +
  labs(color = "Genotype", fill = "Genotype")

ggsave("LarvalSurvival95CI50btsp.png", width=5, height=5)
```