---
title: "vgscKOInsecticideTesting"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(lme4)
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(ecotox)
library(arm)
library(emmeans)
```


## Deltamethrin bottle assays

```{r}
df_Delta <- read.csv("DeltaDeadAlive.csv")
head(df_Delta)
```

GLM with probit link
```{r}
Delta <- glmer(cbind(Dead,Alive) ~ log10(Dose) * Genotype + (1 | Dose:Bottle),
              data = df_Delta,
              family = binomial(link = "probit"))

summary(Delta)
```

Check for overdispersion

```{r}
rp <- residuals(Delta, type = "pearson")
Delta_over <- sum(rp^2)/df.residual(Delta)
Delta_over
```

Check residuals (should be <2 for good fit)

```{r}
Delta_Model_DevResid <- resid (Delta, type = "deviance")
hist(Delta_Model_DevResid)

```

Check variance of residuals across the range of fitted values (most points should fall within the grey line)

```{r}
Delta_Model_Predict <- predict(Delta)
binnedplot(Delta_Model_Predict, Delta_Model_DevResid)
```

Check for outliers (Cook's distance >1 indicates outliers)
```{r}
plot(Delta, which = 4)
```

Model fit looks good

### Pairwise comparison of genotype impact
Overall impact of genotype
```{r}
emmeans(Delta, pairwise ~ Genotype)
```

Impact of genotype at different insecticide levels

```{r}
emmeans(Delta, pairwise ~ Genotype | Dose,
        at = list(Dose = c(0.05, 0.1, 0.5, 1, 5)))
```

**Interpretation**
The vgscKO knockout reduces mortality from low dose deltamethrin
exposure significantly overall, and at all doses except 0.05 ug/ml.

### Produce prediction grid for plotting

```{r}
df_Delta$BottleID <- paste0(format(df_Delta$Dose, trim = TRUE), df_Delta$Bottle)
head(df_Delta)

pred_grid <- expand_grid(
    Dose = sort(unique(df_Delta$Dose)),
    Genotype = unique(df_Delta$Genotype)
) |>
    mutate(logdose = log10(Dose))


head(pred_grid)
```

```{r}
pred_link <- predict(
  Delta, newdata = pred_grid, type = "link",
  se.fit = TRUE, re.form = NA
)

inv_link <- pnorm

pred_df <- pred_grid |>
    mutate(
        fit_link = pred_link$fit,
        se_link = pred_link$se.fit,
        fit = inv_link(fit_link),
        lwr = inv_link(fit_link - 1.96 * se_link),
        upr = inv_link(fit_link + 1.96 * se_link)
        )

# Aggregate observed proportions
obs_df <- df_Delta |>
    group_by(Dose, Genotype) |>
    summarise(Dead = sum(Dead),Total = sum(Total), prop = Dead/Total, .groups = "drop")
```

Plot

```{r}
genotype_cols2 <- c(
  "vgscKO" = "#0B3C8C",
  "WT" = "#B11226"
)

genotype_names <- c(
  "vgscKO" = expression(vgscKO^"+"),
  "WT" = expression(vgscKO^"-")
)

ggplot() +
    geom_ribbon(data = pred_df, aes(Dose, ymin = lwr, ymax = upr, fill = Genotype), alpha = 0.18) +
    geom_line(data = pred_df, aes(Dose, y = fit, colour = Genotype), linewidth = 1) +
    geom_point(data = obs_df, aes(Dose, y = prop, colour = Genotype), size = 2.5) +
    scale_color_manual(values = genotype_cols2, labels = genotype_names) +
    scale_fill_manual(values = genotype_cols2, labels = genotype_names) +
    scale_x_log10(breaks = sort(unique(df_Delta$Dose))) +
    scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0,1)) +
    labs(x = "Deltamethrin dose (µg/ml)", y = "Mortality (%)") +
    theme_classic() +
    theme(legend.position = "bottom")
```


## DDT

Load data

```{r}
df_DDT <- read.csv("DDTDeadAlive.csv")
head(df_DDT)
```

GLM with probit link

```{r}
DDT <- glmer(cbind(Dead,Alive) ~ log10(Dose) * Genotype + (1 | Dose:Bottle),
              data = df_DDT,
              family = binomial(link = "probit"))

summary(DDT)
```

Check for overdispersion

```{r}
rp <- residuals(DDT, type = "pearson")
DDT_over <- sum(rp^2)/df.residual(DDT)
DDT_over
```

Check residuals (should be <2 for good fit)

```{r}
DDT_Model_DevResid <- resid (DDT, type = "deviance")
hist(DDT_Model_DevResid)

```

Check variance of residuals across the range of fitted values (most points should fall within the grey line)

```{r}
DDT_Model_Predict <- predict(DDT)
binnedplot(DDT_Model_Predict, (DDT_Model_DevResid))
```

Check for outliers (Cook's distance >1 indicates outliers)
```{r}
plot(DDT, which = 4)
```

Model fit looks good

### Pairwise comparison of genotype impact
Overall impact of genotype
```{r}
emmeans(DDT, pairwise ~ Genotype)
```

Impact of genotype at different insecticide levels

```{r}
emmeans(DDT, pairwise ~ Genotype | Dose,
        at = list(Dose = c(0.5, 1, 5, 10, 25, 50, 100)))
```

**Interpretation**
The vgscKO knockout reduces mortality from low dose DDT exposure
significantly overall, and at all doses except 0.5 ug/ml and 1 ug/ml.


### Produce prediction grid for plotting

```{r}
df_DDT$BottleID <- paste0(format(df_DDT$Dose, trim = TRUE), df_DDT$Bottle)
head(df_DDT)

pred_grid <- expand_grid(
    Dose = sort(unique(df_DDT$Dose)),
    Genotype = unique(df_DDT$Genotype)
) |>
    mutate(logdose = log10(Dose))


head(pred_grid)
```

```{r}
pred_link <- predict(
  DDT, newdata = pred_grid, type = "link",
  se.fit = TRUE, re.form = NA
)

inv_link <- pnorm

pred_df <- pred_grid |>
    mutate(
        fit_link = pred_link$fit,
        se_link = pred_link$se.fit,
        fit = inv_link(fit_link),
        lwr = inv_link(fit_link - 1.96 * se_link),
        upr = inv_link(fit_link + 1.96 * se_link)
        )

# Aggregate observed proportions
obs_df <- df_DDT |>
    group_by(Dose, Genotype) |>
    summarise(Dead = sum(Dead),Total = sum(Total), prop = Dead/Total, .groups = "drop")

```

Plot

```{r}
genotype_cols2 <- c(
  "vgscKO" = "#0B3C8C",
  "WT" = "#B11226"
)

genotype_names <- c(
  "vgscKO" = expression(vgscKO^"+"),
  "WT" = expression(vgscKO^"-")
)

ggplot() +
    geom_ribbon(data = pred_df, aes(Dose, ymin = lwr, ymax = upr, fill = Genotype), alpha = 0.18) +
    geom_line(data = pred_df, aes(Dose, y = fit, colour = Genotype), linewidth = 1) +
    geom_point(data = obs_df, aes(Dose, y = prop, colour = Genotype), size = 2.5) +
    scale_color_manual(values = genotype_cols2, labels = genotype_names) +
    scale_fill_manual(values = genotype_cols2, labels = genotype_names) +
    scale_x_log10(breaks = sort(unique(df_DDT$Dose))) +
    scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0,1)) +
    labs(x = "DDT dose (µg/ml)", y = "Mortality (%)") +
    theme_classic() +
    theme(legend.position = "bottom")
```

## LC tables for both insecticides

Load combined data

```{r}
df <- read.csv("InsecticideDeadAlive.csv")
head(df)
```

```{r}
df_clean <- subset(df,
                   Dose > 0 &
                   !is.na(Dose) &
                   !is.na(Dead) &
                   !is.na(Total) &
                   Total > 0
)
nrow(df)
nrow(df_clean)
```

```{r}
KO_delta <- LC_probit((Dead/Total) ~ log10(Dose),
          p = c(50,90,95,99),
          weights = Total,
          data = df_clean,
          subset = df_clean$Genotype == "vgscKO" & df_clean$Insecticide == "Deltamethrin"
)

WT_delta <- LC_probit((Dead/Total) ~ log10(Dose),
          p = c(50,90,95,99),
          weights = Total,
          data = df_clean,
          subset = df_clean$Genotype == "WT" & df_clean$Insecticide == "Deltamethrin"
)

KO_DDT<- LC_probit((Dead/Total) ~ log10(Dose),
          p = c(50,90,95,99),
          weights = Total,
          data = df_clean,
          subset = df_clean$Genotype == "vgscKO" & df_clean$Insecticide == "DDT"
)

WT_DDT<- LC_probit((Dead/Total) ~ log10(Dose),
          p = c(50,90,95,99),
          weights = Total,
          data = df_clean,
          subset = df_clean$Genotype == "WT" & df_clean$Insecticide == "DDT"
)

```

```{r}
KO_delta2 <- KO_delta %>%
  mutate(Genotype = "vgscKO",
         Insecticide = "Deltamethrin")

WT_delta2 <- WT_delta %>%
  mutate(Genotype = "WT",
         Insecticide = "Deltamethrin")

KO_DDT2 <- KO_DDT %>%
  mutate(Genotype = "vgscKO",
         Insecticide = "DDT")

WT_DDT2 <- WT_DDT %>%
  mutate(Genotype = "WT",
         Insecticide = "DDT")

LC_all <- bind_rows(KO_delta2, WT_delta2, KO_DDT2, WT_DDT2)

LC_trimmed <- LC_all %>%
  dplyr::select(Genotype, Insecticide, p, dose, LCL, UCL, se)

LC_trimmed
```
