---
title: "vgscKO fecundity analysis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(emmeans); packageVersion("emmeans")
library(glmmTMB); packageVersion("glmmTMB")
library(MASS); packageVersion("MASS")
library(arm); packageVersion("arm")
library(DHARMa); packageVersion("DHARMa")
```

Loading data

```{r}
df <- read.csv("HatchData.csv")
head(df)
```

Removing batches with fewer than 10 larvae

```{r}
df_clean <- na.omit(df)
df_clean <- subset(df_clean, Larvae > 10)

df_clean$Genotype <- factor(df_clean$Genotype, levels = c("WT", "vgscKO"))
df_clean$vgscKOParent <- factor(df_clean$vgscKOParent, levels = c("female", "male"))
```

### Egg laying

Negative binomial GLM to determine impact of Genotype and vgscKOParent on number of eggs laid

```{r}
m_eggs <- glm.nb(Eggs ~ Genotype * vgscKOParent, data = df_clean)
summary(m_eggs)
```

```{r}
m_eggs_no_int <- glm.nb(Eggs ~ Genotype + vgscKOParent, data = df_clean)
anova(m_eggs_no_int, m_eggs, test = "Chisq")
```

Model with interaction term kept

Check for overdispersion

```{r}
dispersion_eggs <- sum(residuals(m_eggs, type = "pearson")^2) / df.residual(m_eggs)
dispersion_eggs
```


Check residuals (should be <2 for good fit)

```{r}
m_eggs_Model_DevResid <- resid (m_eggs, type = "deviance")
hist(m_eggs_Model_DevResid)
```

Check variance of residuals across the range of fitted values (most points should fall within the grey line)

```{r}
m_eggs_Model_Predict <- predict(m_eggs)
binnedplot(m_eggs_Model_Predict, (m_eggs_Model_DevResid))
```

Check for outliers (Cook's distance >1 indicates outliers)
```{r}
plot(m_eggs, which = 4)
```

Impact of genotype on egg production

```{r}
pairs(emmeans(m_eggs, ~ Genotype, weights = "proportional"))
```

**Interpretation**

vgscKO females produce slightly larger egg batches than vgscKO males - but there is no 
overall impact of the vgscKO knock out on egg production.


### Larvae numbers

Negative binomial GLM for larvae numbers, with an offset factor for egg number.

```{r}
m_larvae_eff <- glm.nb(
  Larvae ~ Genotype * vgscKOParent + offset(log(Eggs)),
  data = df_clean
)
summary(m_larvae_eff)
```

```{r}
dispersion_larvae <- sum(residuals(m_larvae_eff, type = "pearson")^2) / df.residual(m_larvae_eff)
dispersion_larvae
```


Check residuals (should be <2 for good fit)

```{r}
larvae_Model_DevResid <- resid (m_larvae_eff, type = "deviance")
hist(larvae_Model_DevResid)
```

Slightly skewed but still relatively normal distribution around 0

Check variance of residuals across the range of fitted values (most points should fall within the grey line)

```{r}
larvae_Model_Predict <- predict(m_larvae_eff)
binnedplot(larvae_Model_Predict, (larvae_Model_DevResid))
```

Check for outliers (Cook's distance >1 indicates outliers)
```{r}
plot(m_larvae_eff, which = 4)
```

Double-check with DHARMa
```{r}
sim <- simulateResiduals(m_larvae_eff)
plot(sim)
testDispersion(sim)
```

Model fits well

Estimated marginal means 

```{r}
pairs(emmeans(m_larvae_eff, ~ Genotype, type = "response"))
pairs(emmeans(m_larvae_eff, ~ vgscKOParent, type = "response"))
```

**Interpretation**

No significant impact of genotype or vgscKO parent on larval output.

### Hatch rate analysis

Quasibinomial GLM

```{r}
df_hatch <- subset(df_clean, Eggs > 0 & Larvae <= Eggs)

m_hatch_q <- glm(
  cbind(Larvae, Eggs - Larvae) ~ Genotype * vgscKOParent,
  family = quasibinomial,
  data = df_hatch
)
summary(m_hatch_q)
```

```{r}
dispersion_hatch <- sum(residuals(m_hatch_q, type = "pearson")^2) / df.residual(m_hatch_q)
dispersion_hatch
```

Dispersion far too high - try a beta-binomial model

```{r}
library(glmmTMB)
m_hatch_bb <- glmmTMB(
  cbind(Larvae, Eggs - Larvae) ~ Genotype * vgscKOParent,
  family = betabinomial(link = "logit"),
  data = df_hatch
)
summary(m_hatch_bb)
```

**Check for model fit using DHARMa**

Overdispersion

```{r}
sim <- simulateResiduals(m_hatch_bb)
plot(sim)
testDispersion(sim)
```


Check for outliers
```{r}
testOutliers(sim)
```


Beta-binomial model fits well


Overall (marginal) genotype effect:
```{r}
pairs(emmeans(m_hatch_bb, ~ Genotype, type = "response"))
pairs(emmeans(m_hatch_bb, ~ vgscKOParent, type = "response"))
```

**Interpretation**

There was no significant effect of Genotype or vgscKO Parent sex 
on the hatch rate of eggs.